Calendar-based graphics for visualising people's daily schedules

### 2017-07-19 main.tex
spelling
"proof reading by Stuart. Many thanks to Stuart.

\Abstract{
This paper describes a \code{frame\_calendar} function that organises
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Facetting by the usual features such as day and month, was not
sufficient to examine the behaviour. Making displays on a monthly
calendar format helps to understand pedestrian patterns relative to
events such as work days, weekends, holidays, and special events. The
layout algorithm has several format variations and options. It is
implemented in the R package \pkg{sugrrants}.
}

@@

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. A small multiples, shown in Figure
\ref{fig:time-series-plot}, gives an overall picture of foot traffic at
a selection of 3 sensors. Figure \ref{fig:facet-time} provides another
aspect of temporal patterns in the more detailed level by faceting day
of the week. It lends itself to a number of issues that make exploratory
data visualisation challenging in many temporal-context applications
involving human behaviours:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily result from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurred events).
\item
  Since the data are often collected at daily frequency or a time scale
  more frequent than daily, they typically involve a large number of
  observations spanning over a long time period.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need in comparing and contrasting
  between locations.
\end{enumerate}

@@

\caption[Time series plot about the number of pedestrians over the year of 2016 at a selection of 3 sensors in the city of Melbourne]{Time series plot about the number of pedestrians over the year of 2016 at a selection of 3 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distincitve across these 3 sensors. There is an eye-catching spike occurred to State Library, which is caused by the annual event--White Night--on 20th of Feburary.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

A collection of data plots, organised in a familiar format, offer a
useful and intuitive way to tell a richer story and reveal more complex
relationships. \citet{Wickham2012glyph} embedded a sequence of daily
temperature line charts into a glyph map according to the spatial
locations; and \citet{R-geofacet} provided methods to arrange the data
plots of many types into a preserved geographical grid in the
\pkg{geofacet} package. They both attempt to have the self-contained
data plots in spatial context.

Alternatively, calendar-based graphics turn out to be a useful tool in
unfolding human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
weekdays and weekends, highlights public holiday, and presents other
known seasons like school vacations, all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in a couple of R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are too
constrained to colour-encoding graphics and day of week as the smallest
time scale. Time of day, which serves as one of the most important
aspect in explaining variations arisen from pedestrian sensor data, will
be neglected through daily aggregation. Additionally if simply using
coloured blocks rather than curves, it may become perceptually difficult
to estimate the shape positions and changes, although using the curves
comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plot
consisting of multiple time series. The proposed algorithm has been
implemented to the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics
\citep{wilkinson2006grammar, wickham2010layered}.

@@

\caption[The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs]{The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of 3 by 4 on the monthly calendar format presents all the traffic in 2016. The disparities between weekday and weekend along with public holiday pop out to viewers in the clearer manner.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar, requires cells
for days, embedded in blocks corresponding to months, organised into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5. These cells provide a micro canvas on which to plot
the data. The first day of the month could be any of Monday-Sunday,
which is determined by the year of the calendar. Months are of different
length days, ranging form 28-31, and each month could extend over six
weeks but the convention in these months is to wrap the last few days up
to the top row of the block. The notation for creating these cells is as
follows:

@@

Each cell forms a canvas on which to draw the data. Consider the canvas
to have limits \([0, 1]\) horizontally and vertically. For the
pedestrian sensor data, within each cell hour is plotted horizontally
and count is plotted vertically. Each variable is scaled to have values
between \([0,1]\), using the minimum and maximum of all the data values
to be displayed assuming of fixed scales. Let \(h\) be the scaled hour,
and \(c\) the scaled count.

@@

Note that for the vertical direction, the top left is the starting point
of the grid, hence the subtractions, and resulting negative values to
lay out the cells. Within each cell, the starting position is bottom
left.

The algorithm can be relatively easily extended to layout from a single
month to a few years, based on the period of time to be visualised.
\((M, N)\) can be determined by the user using the arguments of
\code{nrow} and \code{ncol}. If the one would like to visualise data
spanning over three years, for example, \(M = 12\) and \(N = 3\) seem to
be an appropriate choice when assessing the differences of a given month
across the years.

We only illustrate that grids are laid out over the horizontal direction
in the paper. The vertical direction can be enabled by swapping \(i\)
and \(j\) in the algorithm stated above when the argument \code{dir} is
set to \code{"v"}. It particularly benefits those users who get
accustomed to calendars of vertical organisation in some countries.

@@

There are a few options provided for the \code{frame_calendar} function
to adjust the display and they are shown as follows:

@@

Assuming that tidy data \citep{wickham2014tidy} is the underlining data
structure, the \code{x} takes a variable that will be mapped to the x
axis and the \code{y} mapped to the y axis for the later plotting. In
Figure \ref{fig:flinders-2016}, for example, the \code{x} is the
variable specifying the time of the day, and the \code{y} is the
variable suggesting the hourly counts. The \code{date} argument is given
by the date variable that determines the correct order in the calendar
layout. We shall describe some of the arguments that allow for different
displays.

@@

The algorithm described in Section \ref{sec:algorithm} is for the most
common calendar layout--``monthly'' calendar, and it can be simplified
to accommodate the other two types of calendar formats. One is comprised
of days of a week in columns and weeks of a year in rows, and the other
is days of a month in columns and months of a year in rows, which we
refer to as ``weekly'' and ``daily'' calendar respectively. Which layout
to be used is controlled by the \code{calendar} argument. Due to the way
of the arrangement, the weekly calendar puts more emphases on days of a
week over days of a year, whereas the daily one serves as the opposite.
The monthly format can be considered as the advanced twist of both
weekly and daily. Temporal patterns lead to which format to be employed.
The weekly calendar can be a nice attempt if the most variations can be
characterised by days of a week. On the other hand, the absence of
weekly effect but the presence of yearly effect can direct to the daily
calendar. When both are present, the monthly calendar appears to be a
better choice.

@@

When \code{polar = TRUE}, the polar transformation is carried out for
the data. Again the computation remain similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} is considered
as the spiral display of Figure \ref{fig:flinders-2016}.

@@

Section \ref{sec:algorithm} discusses the implementation applied to the
fixed scale which is using the whole range of all the data values. The
\code{scale} argument controls the scaling of the display. The fixed
scale (\code{fixed}) is the default, meaning the data values in all
positions to be scaled. The remaining options include free scale within
each cell (\code{free}), cells derived from the same day of the week
(\code{free_wday}), or cells from the same day of the month
(\code{free_mday}). It utilises the comparisons of absolute or relative
values, and the emphases of various types of variations.

Grouping the cells based on the same time period gives rise to the
different scales. For example, the minimums and maximums obtained from
each cell, that is every day \((i, j)\) together with \((m, n)\), result
in local scale. The overall variation gives way to the individual shape.
Figure \ref{fig:flinders-free} is an example of plotting line charts
scaled locally. The daily variation is made more distinctive, compared
to Figure \ref{fig:flinders-2016}.

@@

\caption[Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days]{Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive at the expense of the magnitude comparison loss.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}

\subsection{Reference lines and
labels}\label{reference-lines-and-labels}

>>>>

\Abstract{
This paper describes a \code{frame\_calendar} function that organises
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Facetting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format variations and options. It is implemented in the R
package \pkg{sugrrants}.
}

@@

% Pandoc header

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 sensors installed that
count pedestrians every hour across the downtown area up until the end
of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. A small multiples, shown in Figure
\ref{fig:time-series-plot}, gives an overall picture of the foot traffic
at 3 different sensors. Figure \ref{fig:facet-time} provides more
details of the temporal patterns by facetting on day of the week. The
sensor data, like many temporal datasets on human behaviours, lends
itself to a number of exploratory data visualisation challenges:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Variations primarily result from multiple time scales including time
  of day, day of week, and day of year (such as public holiday and
  recurring events).
\item
  Since the data are often collected at sub-daily frequencies, they
  typically involve a large number of observations.
\item
  Measurements of a single type are made at multiple locations at a
  given time point, which creates the need for comparing and contrasting
  between locations.
\end{enumerate}

@@

\caption[Time series plot about the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne]{Time series plot about the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distincitve across these 3 sensors. There is an eye-catching spike which occurred at State Library, caused by the annual event--White Night--on 20th of Feburary.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

A collection of data plots, organised in a familiar format, offer a
useful and intuitive way to understand complex data.
\citet{Wickham2012glyph} embedded a sequence of daily temperature line
charts into a glyph map according to the spatial locations; and
\citet{R-geofacet} provided methods in the \pkg{geofacet} package to
arrange data plots into a grid, while preserving the geographical
layout. They both attempt to enable one to see the spatial context of
each self-contained data plot.

Alternatively, calendar-based graphics turn out to be a useful tool in
unfolding human-related activities over time. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
weekdays and weekends, highlights public holidays, and presents other
known seasons like school vacations; all of which have influence over
the turn-outs in the office. The calendar-based heatmap was implemented
in two R packages: \pkg{ggTimeSeries} \citep{R-ggTimeSeries} and
\pkg{ggcal} \citep{R-ggcal}. However, these techniques are limited to
colour-encoding graphics and are unable to use time scales smaller than
day. Time of day, which serves as one of the most important aspects in
explaining variations arisen from pedestrian sensor data, will be
neglected through daily aggregation. Additionally if simply using
coloured blocks rather than curves, it may become perceptually difficult
to estimate the shape positions and changes, although using the curves
comes with the cost of more display capacity
\citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm via linear algebra tools to go beyond the
calendar-based heatmap. The approach is developed with these conditions
in mind: (1) to make time of day present in addition to the existing
temporal components such as day of week and day of year, (2) to
incorporate line graphs and other types of glyphs into the graphical
toolkit for the calendar layout, (3) to enable an overlaying plot
consisting of multiple time series. The proposed algorithm has been
implemented in the \code{frame_calendar} function in the \pkg{sugrrants}
package \citep{R-sugrrants} using \proglang{R} \citep{R-base}.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is an
implementation of the grammar of graphics
\citep{wilkinson2006grammar, wickham2010layered}.

@@

\caption[The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs]{The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a 3 by 4 monthly grid represents all the traffic in 2016. The disparities between weekday and weekend along with public holiday are immediately apparent.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar, requires cells
for days, embedded in blocks corresponding to months, organised into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5. These cells provide a micro canvas on which to plot
the data. The first day of the month could be any of Monday-Sunday,
which is determined by the year of the calendar. Months are of different
length days, ranging from 28-31, and each month could extend over six
weeks but the convention in these months is to wrap the last few days up
to the top row of the block. The notation for creating these cells is as
follows:

@@

Each cell forms a canvas on which to draw the data. Consider the canvas
to have limits \([0, 1]\) horizontally and vertically. For the
pedestrian sensor data, within each cell, hour is plotted horizontally
and count is plotted vertically. Each variable is scaled to have values
in \([0,1]\), using the minimum and maximum of all the data values to be
displayed, assuming fixed scales. Let \(h\) be the scaled hour, and
\(c\) the scaled count.

@@

Note that for the vertical direction, the top left is the starting point
of the grid which is why subtraction is performed. The resulting
negative values place the cells down the vertical direction. Within each
cell, the starting position is the bottom left.

@@

There are several options provided for the \code{frame_calendar}
function to initialize and adjust the display of a calander plot:

@@

Assuming that tidy data \citep{wickham2014tidy} is the underlying data
structure, the parameter \code{x} takes a variable that will be mapped
to the x axis and the parameter \code{y} mapped to the y axis for plot
construction. In Figure \ref{fig:flinders-2016}, for example, the
\code{x} is the variable specifying the time of the day, and the
\code{y} is the variable representing the hourly counts. The \code{date}
argument is given by the date variable that determines the correct order
of the calendar layout.

The algorithm can be extended to display data from a single month up to
a few years. The number of rows and columns in the layout can be
specified with the arguments \code{nrow} and \code{ncol}. If the one
would like to visualise data spanning over three years, for example,
\code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when
assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid
out horizontally. The vertical direction can be enabled by swapping
\(i\) and \(j\) in equation \ref{eq:grid}, which occurs when the
argument \code{dir} is set to \code{"v"}. This benefits users who are
accustomed to calendars that are organised vertically, which is common
in some countries. Next we shall describe some of the arguments that
allow for alternate displays.

@@

The algorithm described in Section \ref{sec:algorithm} is for the most
common calendar layout, the ``monthly'' calendar, but it can be
simplified to accommodate the other two types of calendar formats. One
is comprised of days of a week in columns and weeks of a year in rows,
and the other is days of a month in columns and months of a year in
rows, which we refer to as the ``weekly'' and ``daily'' calendar,
respectively. The layout to be used is controlled by the \code{calendar}
argument. The weekly calendar puts more emphasis on days of a week over
days of a year, whereas the daily calender does the opposite. The
monthly calender can be considered as the advanced variant of both
weekly and daily. Temporal patterns motivate which variant should be
employed. The weekly calendar is appropriate if most of the variation
can be characterised by days of the week. On the other hand, the daily
calendar should be used when there is a yearly effect but not a weekly
effect in the data. When both effects are present, the monthly calendar
would be a better choice.

@@

When \code{polar = TRUE}, the polar transformation is carried out on the
data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} is considered
as the spiral display of Figure \ref{fig:flinders-2016}.

@@

Section \ref{sec:algorithm} discusses the implementation applied to a
fixed scale, which uses the range over all data values. The \code{scale}
argument controls the scaling of the display. The fixed scale
(\code{fixed}) is the default, meaning that the data values in all
positions are scaled. The remaining options include: free scale within
each cell (\code{free}), cells derived from the same day of the week
(\code{free_wday}), or cells from the same day of the month
(\code{free_mday}). The scaling allows for the comparisons of absolute
or relative values, and the emphasis of different temporal variations.

Grouping the cells based on the same time period gives rise to the
different scales. For example, the minimums and maximums obtained from
each cell produce a local scale. The overall variation gives way to the
individual shape. Figure \ref{fig:flinders-free} is an example of
plotting line charts scaled locally. The daily variation is more
distinctive compared to Figure \ref{fig:flinders-2016}.

@@

\caption[Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days]{Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is more difficult to compare the size of peaks between days.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}
